---
title: "Documentation log"
output:
  md_document:
    variant: gfm
    preserve_yaml: true
layout: default
nav_order: 6
math: mathjax
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.path='../assets/images/')
```

# Documentation log count

Last update:
```{r, echo=FALSE, message=FALSE, warning=FALSE}
print(Sys.Date())
```

This doc was built with: `rmarkdown::render("documentation_log.Rmd", output_file = "../pages/documentation_log.md")`

Below is the latest count of documentation pages. 
This figure is updated sporadically to track the current page count and estimated growth rate.
This count only considers the main sources in `./pages` but ignores other code pages, images, etc. 


```{r word-count,  echo=FALSE, message=FALSE, warning=FALSE}
library(purrr)
library(lubridate)
library(ggplot2)
library(dplyr)
library(broom)
library(patchwork)

# Set the path to the directory containing the Markdown files
directory_path <- "../pages/"

# List all Markdown files in the directory
md_files <- list.files(path = directory_path, pattern = "\\.md$", full.names = TRUE)

# Define a function to count lines and words in a file
count_lines_words <- function(file_path) {
  lines <- readLines(file_path)
  word_count <- sum(sapply(lines, function(line) length(strsplit(line, "\\s+")[[1]])))
  line_count <- length(lines)
  data.frame(FileName = basename(file_path), LineCount = line_count, WordCount = word_count)
}

# Apply the function to each Markdown file and bind the results into a single data frame
results <- map_df(md_files, count_lines_words)
# sum(results$LineCount)
# sum(results$WordCount)
```


```{r file_count_plot, echo=FALSE, message=FALSE, warning=FALSE}

# Load the data from the log file
df <- read.table(file = "./file_counts/file_count_log.md",
                 header = FALSE,
                 col.names = c("Date", "FileCount"))

# Convert the date from 'YYYYMMDD' to a Date object
df$Date <- ymd(df$Date)

# Convert Date to numeric for regression
df$DateNumeric <- as.numeric(df$Date)

# Linear regression to estimate growth rate
fit <- lm(FileCount ~ DateNumeric, data = df)
# summary(fit)  # Show summary of regression

# Extracting the estimated growth rate per day
growth_rate <- coef(fit)["DateNumeric"]  # Slope coefficient
annual_growth_rate <- growth_rate * 365  # Convert daily rate to annual rate

# Plotting the data with the regression line
p1 <- df %>%
  ggplot(aes(x = Date, y = FileCount)) +
  geom_point() +  # Plot points
  geom_line() +   # Connect points with line
   geom_smooth(method="lm", color="grey", alpha=0.2)  +  # Regression line
  theme_bw() +    # Use a minimalistic theme
  labs(title = "Number of documentation pages over time",
       x = "Date",
       y = "Number of independent pages",
       subtitle = paste0("Estimated annual growth rate: ", round(annual_growth_rate, 2), "pages/year", 
                         "\nCurrent page count: ", max(df$FileCount),
                         "\nTotal line count: ", sum(results$LineCount),
                         "\nTotal word count: ", sum(results$WordCount)
))  # Add labels and subtitle

# Display the plot
print(p1)

# Display regression summary as part of the document output
cat("Regression Summary:\n")
print(summary(fit))

# Print the estimated annual growth rate
cat("Estimated Annual Growth Rate (pages/year):", round(annual_growth_rate, 2))
cat("Current count:", max(df$FileCount))
```


```{r word-count-plots,  echo=FALSE, message=FALSE, warning=FALSE}
# Plotting Line Count Histogram
p1 <- ggplot(results, aes(x = LineCount)) +
  geom_histogram( color = "black") +
  labs(title = "Histogram of line counts", x = "Line lount", y = "Frequency") +
  theme_minimal()

# Plotting Word Count Histogram
p2 <- ggplot(results, aes(x = WordCount)) +
  geom_histogram(color = "black") +
  labs(title = "Histogram of word counts", x = "Word count", y = "Frequency") +
  theme_minimal()

# Print the plots
print(p1 + p2)
```



## Change log

See `generate_gitlog.sh` for usage.

* [View Git Log 2025](../gitlog_2025.txt)
* [View Git Log 2024](../gitlog_2024.txt)
* [View Git Log 2023](../gitlog_2023.txt)

(current year log may not be up-to-date)

```{r echo=FALSE, results='asis'}
cat("## 2025", sep="\n")
gitlog_path <- "../gitlog_2025.txt"  # Adjust the path as needed

# Check if the file exists
if (file.exists(gitlog_path)) {
  # Read and print the contents of the git log
  log_contents <- readLines(gitlog_path)
  cat(log_contents, sep="\n\n")
} else {
  cat("The git log file for 2025 is not available.")
}
```

```{r echo=FALSE, results='asis'}
cat("## 2024", sep="\n")
gitlog_path <- "../gitlog_2024.txt"
log_contents <- readLines(gitlog_path)
cat(log_contents, sep="\n\n")
```

```{r echo=FALSE, results='asis'}
cat("## 2023", sep="\n")
gitlog_path <- "../gitlog_2023.txt"
log_contents <- readLines(gitlog_path)
cat(log_contents, sep="\n\n")
```
